Introducing statistics with \texttt{intRo}

### 2015-02-16 student-paper-competition-2014/paper.Rmd
homeworks to be done using intRo

---
title: "Introducing statistics with \\texttt{intRo}"
author: "Eric Hare & Andee Kaplan"
date: "December 14, 2014"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
bibliography: refs.bib
geometry: margin=0.8in
---
\thispagestyle{empty}

\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper we introduce the application and explore the design decisions underlying \texttt{intRo} as well as highlight some challenges and advantages of reactive programming.
\end{abstract}

\twocolumn
\setcounter{page}{1}

# Introduction

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been a great development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions.

In teaching STAT 226 at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class. Scrimshaw notes in his manuscript that "open-ended packages, like any others, may create obstacles to learning simply through their lack of user-friendliness in the sheer mechanics of operating them, rather than any intrinsic difficulty in the content...." [@scrimshaw2001computers] In our own experience teaching, students' struggles were often directly related to the use of the software and not any sort of fundamental misunderstanding of the material, in agreement with Scrimshaw's finding.

Other software has recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning \texttt{R} within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easy for an R-based analysis to be converted to an interactive web application. This in turn, led to the creation of an introductory statistics application which we call \texttt{intRo}, available at http://www.intro-stats.com. \texttt{intRo} offers a number of key advantages over traditional statistics software, which will be discussed at length in this text.

@@

## Easy

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{Typical user experience of generating a result in \texttt{intRo}. In this instance, a mosaic plot (and its corresponding code) was created and saved to the final compendium.}
\label{fig:user_experience}
\end{figure}

In order for an application to be useful to students in their first statistics course, it can't hog the spotlight. Statistics as a subject should shine through as a fun and interesting topic to study, rather than being bogged down by difficult to learn software. With this in mind, in the development of \texttt{intRo} we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find exactly what they need easily. With this easy to use framework, we also made the functionality available in \texttt{intRo} the minimal necessary for an introductory statistics course. Additional functionality is nice, but if it clutters the UI beyond the point of usability, it will frustrate an introductory student. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}, and view the code producing the result.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}

@@

There are multiple modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics - and more modules can be easy added by an instructor, as detailed in section \ref{sec:extensible}. The default modules support uploading and downloading a dataset, transforming variables, graphical and numerical summaries, simple linear regression, contingency tables, and T-tests.

In addition to making \texttt{intRo} intuitive to use, we've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo/ and hyperlinked within the application itself. This documentation covers all default modules and makes learning to use \texttt{intRo} even more painless.

@@

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}.}
\label{fig:app_creation_modules}
\end{figure*}

@@

\footnotesize
```{r ui, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=25)}
## Source ui
module_info <- read.table("modules/modules.txt", header = TRUE, sep=",")
sapply(file.path("modules", dir("modules")[dir("modules") != "modules.txt"], "ui.R"), source)

## mylist is a list containing the different ui module code
## Create the UI
shinyUI(
    navbarPage("intRo", id="top-nav",  theme = "bootstrap.min.css",
               tabPanel(title="", icon=icon("home"),
                        fluidRow(
                            do.call(navlistPanel, c(list(id = "side-nav", widths = c(2, 10)), mylist))
                        )
               ), ...
    ))
```
\normalsize

@@

\footnotesize
```{r server, eval=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=25)}
shinyServer(function(input, output, session) {
  ## Module info
  module_info <- read.table("modules/modules.txt", header = TRUE, sep=",")

@@

  ## Source the modules
  for (mod in modules_tosource) {
      source(mod, local = TRUE)
  }
})
```
\normalsize
In this way, we were able to have \texttt{intRo} be fully extensible, its structure and functionality dependent entirely on the modules present within the application.

@@

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in \texttt{intRo}'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}
Part two of this text discusses probability and sampling distributions, which is outside the current realm of \texttt{intRo}, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in \texttt{intRo}. Finally, part four of the text discusses the association between variables, including regression, functions that \texttt{intRo} performs with ease. Some of the more advanced topics of this section are currently unsupported by \texttt{intRo}, including multiple regression and nonparametric statistics.

@@

In addition to the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more useful to both students and instructors. In particular, we have begun development on an \texttt{R} package which will allow \texttt{intRo} modules to be created automatically from user written \texttt{R} code. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo} as well as translate user code to \texttt{intRo} compatible code as populates the necessary files. This will vastly improve \texttt{intRo}'s flexibility and allow it to be used in a wider range of curricula, including more advanced statistics courses. Furthermore, we would like to utilize the expanded interactive capabilities of \texttt{ggvis} in order to make \texttt{intRo's} plots more engaging to students. One way to do this would be implementing linked plots, in which interactions with one plot are reflected in other plots that illustrate the same data. This would be particularly useful in the regression module so that students could explore observations with high influence and leverage.

@@

Challenges do exist with regards to the wider adoption of \texttt{intRo}. For instance, we will need to monitor how well the server hosting \texttt{intRo} handles the load of dozens of students performing data analyses at once. If performance issues are encountered, the infrastructure used may need to be expanded to handle current and future load. An unknown quantity will be how feasible it is to increase adoption of \texttt{intRo} across the department, as well as to other universities. One limitation of \texttt{intRo} is that uploading a dataset beyond about 30,000 rows tends to be slow. Even once the data is successfully uploaded, the default modules produce results more slowly than with smaller datasets. This is a limitation that should be further investigated if and when \texttt{intRo} sees wider adoption.

>>>>

---
title: "Introducing statistics with \texttt{intRo}"
author: "Eric Hare & Andee Kaplan"
date: "December 14, 2014"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
  html_document: default
bibliography: refs.bib
---
\thispagestyle{empty}
\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper we introduce the application and explore the design decisions underlying \texttt{intRo} as well as highlight some challenges and advantages of reactive programming.
\end{abstract}

\newpage
\setcounter{page}{1}

# Introduction
The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been a great development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions. In teaching STAT 226 at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class. Often, these struggles were directly related to the use of the software and not any sort of fundamental misunderstanding of the material.

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easy for an R-based analysis to be converted to an interactive web application. This in turn, led to the creation of an introductory statistics application which we call \texttt{intRo}.

\texttt{intRo} offers a number of key advantages over traditional statistics software, which will be discussed at length in this text.

\ak{[Add some references about software discouraging people, and mention other new tools out there like swirl.]}

@@

## Easy
In order for an application to be useful to students in their first statistics course, it can't hog the spotlight. Statistics as a subject should shine through as a fun and interesting topic to study, rather than being bogged down by difficult to learn software. With this in mind, in the development of \texttt{intRo} we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find exactly what they need easily. With this easy to use framework, we also made the functionality available in \texttt{intRo} the minimal necessary for an introductory statistics course. Additional functionality is nice, but if it clutters the UI beyond the point of usability, it will frustrate an introductory student.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\textwidth]{images/ui_annotate.png}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure}

@@

There are multiple modules within \texttt{intRo}, which perform specific data analysis tasks that employ an easy to use point-and-click interface. Each module is placed within a higher level category to further help students navigate the tool. The default modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics - and more modules can be easy added by an instructor, as detailed in section \ref{sec:extensible}.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from her computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of \texttt{intRo}. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.

The graphical summaries module provides several common graphical displays of variables in the dataset. First, the user selects the plot desired, and either a single variable or two variables from the dataset, depending on plot type. The variable available to choose will only be those compatible with plot type selected. The plot visible on the right will automatically update when these values are selected. The numeric summaries module is another method of computing summaries of the data. The user is presented with a list of each variable from their dataset. The user can select one or more different variables, and will be presented with a table of some summary statistics. The summary statistics displayed vary depending on the type of the variables. For instance, for a numeric variable, a five number summary will be displayed. Character or factor variables will display counts of each unique value of that particular variable. Additionally, the user has the ability to display summaries by a grouping variable.

Within the set statistics modules, there are three simple statistics tasks that an introductory statistics student will learn in her first semester. The contingency table module allows the user to compute a contingency table for two categorical variables. The results can either be displayed as counts, row percentages, column percentages, or total percentages. The simple linear regression module allows the user to compute a simple linear regression of a dependent variable (y) on an independent variable (x). The module will automatically filter any variables that are non-numeric so that the user may only select variables that are suitable for a simple linear regression. Results of the regression will be displayed on the right. A table of parameter estimates, a scatterplot of the data along with a line of best fit, The R-Squared value, and residual diagnostic plots will all be displayed. A user also has the ability to save residuals and predicted values to their dataset for further analysis. The T test module allows the user to perform either a one variable or two variable Student's T test. The user can select configuration options, including the direction of the test and the confidence level of the test. The results, including the t statistic, degrees of freedom, p-value, and a confidence interval, will be displayed in a text box in the results panel.

In addition to making \texttt{intRo} intuitive to use, we've also created a documentation website that is consistent with the interface of the application and hyperlinked within the application itself. This documentation covers all default modules and makes learning to use \texttt{intRo} even more painless.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{images/app_creation_modules.png}
\caption{This figure depicts how the \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}.}
\label{fig:app_creation_modules}
\end{figure}

@@

\footnotesize
```{r ui, eval=FALSE}
## Source ui
module_info <- read.table("modules/modules.txt", header = TRUE, sep=",")
sapply(file.path("modules", dir("modules")[dir("modules") != "modules.txt"], "ui.R"), source)

## Generates the UI tabs
mylist <- list()
old_heading <- ""
for (i in 1:nrow(module_info)) {
    if (module_info[i,1] != old_heading) {
        if (i != 1) mylist[[length(mylist) + 1]] <- "-----"
        mylist[[length(mylist) + 1]] <- as.character(module_info[i,1])
        old_heading <- module_info[i,1]
    }
    mylist[[length(mylist) + 1]] <- get(paste(module_info[i,2], "ui", sep = "_"))
}

## Create the UI
shinyUI(
    navbarPage("intRo", id="top-nav",  theme = "bootstrap.min.css",
               tabPanel(title="", icon=icon("home"),
                        fluidRow(
                            do.call(navlistPanel, c(list(id = "side-nav", widths = c(2, 10)), mylist))
                        )
               ), ...
    ))
```
\normalsize

@@

\footnotesize
```{r server, eval=FALSE}
shinyServer(function(input, output, session) {
  ## Module info
  module_info <- read.table("modules/modules.txt", header = TRUE, sep=",")

@@

  ## Source the modules
  for (mod in modules_tosource) {
      source(mod, local = TRUE)
  }
}
```
\normalsize
In this way, we were able to have \texttt{intRo} be fully extensible, its structure and functionality dependent entirely on the modules present within the application.

@@

\begin{table}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\label{tab:plottypes}
\caption{The plot and variable types allowed in \texttt{intRo}'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\end{table}
Part two of this text discusses probability and sampling distributions, which is outside the current realm of \texttt{intRo}, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in \texttt{intRo}. Finally, part four of the text discusses the association between variables, including regression, functions that \texttt{intRo} performs with ease. Some of the more advanced topics of this section are currently unsupported by \texttt{intRo}, including multiple regression and nonparametric statistics.

@@

In addition to the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more useful to both students and instructors. In particular, we have begun development on an \texttt{R} package which will allow \texttt{intRo} modules to be created automatically from user written \texttt{R} code. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo} as well as translate user code to \texttt{intRo} compatible code as populates the necessary files. This will vastly improve \texttt{intRo}'s flexibility and allow it to be used in a wider range of curricula, including more advanced statistics courses.

@@

Challenges do exist with regards to the wider adoption of \texttt{intRo}. For instance, we will need to monitor how well the server hosting \texttt{intRo} handles the load of dozens of students performing data analyses at once. If performance issues are encountered, the infrastructure used may need to be expanded to handle current and future load. An unknown quantity will be how feasible it is to increase adoption of \texttt{intRo} across the department, as well as to other universities.
